### 2018-03-17 paper/3-abstractions.tex
Add authors, minor changes

\paragraph{Task description} Any build system requires the user to specify how to
compute the new value for one key, using the (up to date) values of its
dependencies. We call this specification the \emph{task description}.  For
example, in \Excel, the formulae of the spreadsheet constitute the task
description; in \Make the rules in the makefile are the task description.

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up to date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}.

\noindent
Here is a simple build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: Eq k => Build Monad () k v
busy task key store = execState (fetch key) store
  where
    fetch :: k -> State (Store () k v) v
    fetch k = case task fetch k of
        Nothing  -> gets (getValue k)
        Just act -> do v <- act; modify (putValue k v); return v
\end{minted}
\vspace{1mm}

@@

\noindent
Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel (which
uses dynamic dependencies) is perfectly happy. We can express these rules using
our task abstraction as:

@@

% The above definition of correctness needs to be adjusted to accommodate
% non-deterministic tasks and shallow cloud builds; we address this
% in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects}, respectively.

@@

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task key = case task (\k -> Const [k]) key of
                            Nothing         -> []
                            Just (Const ks) -> ks
\end{minted}

@@

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
\hs{runWriterT} to the contents of \hs{Maybe}.
Here is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}:

>>>>

\vspace{-2mm}
\paragraph{Task description} Any build system requires the user to specify how to
compute the new value for one key, using the (up to date) values of its
dependencies. We call this specification the \emph{task description}.  For
example, in \Excel, the formulae of the spreadsheet constitute the task
description; in \Make the rules in the makefile are the task description.

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up to date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: Eq k => Build Monad () k v
busy task key store = execState (fetch key) store
  where
    fetch :: k -> State (Store () k v) v
    fetch k = case task fetch k of
        Nothing  -> gets (getValue k)
        Just act -> do v <- act; modify (putValue k v); return v
\end{minted}
\vspace{1mm}

@@

\noindent
Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel (which
uses dynamic dependencies) is perfectly happy. We can express this spreadsheet
using our task abstraction as:

@@

The presented definition of correctness needs to be adjusted to accommodate
non-deterministic tasks and shallow cloud builds, as will be discussed
in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
respectively.

@@

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task key = case task (\k -> Const [k]) key of
                            Nothing         -> []
                            Just (Const ks) -> ks
\end{minted}
\vspace{1mm}

@@

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
\hs{runWriterT} to the contents of \hs{Maybe}.
Below we give an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

### 2018-03-17 paper/4-build.tex
Add authors, minor changes

\subsubsection{Topological}\label{sec-topological}

@@

\subsubsection{Reordering}\label{sec-reordering}

@@

\subsubsection{Recursive}\label{sec-recursive}

@@

This approach requires that a task may be started, then during that execution
another task will have to be run. Assuming a an IO-driven task structure,
that requires suspending a running task, which can be done with cheap
green threads and blocking (the original approach of \Shake) or using
continuation-passing style (what \Shake does currently). An alternative approach to
suspending a task is to abort it and restart it again later, at the cost
of doing additional work.

@@

\subsubsection{A dirty bit}\label{sec-dirty-bit}

@@

\subsubsection{Verifying traces}\label{sec-verifying-traces}

>>>>

\vspace{-2mm}
\subsubsection{Topological}\label{sec-topological}

@@

\vspace{-2mm}
\subsubsection{Reordering}\label{sec-reordering}

@@

\vspace{-2mm}
\subsubsection{Recursive}\label{sec-recursive}

@@

This approach requires that a task may be started, then during that execution
another task will have to be run. Assuming an IO-driven task structure,
that requires suspending a running task, which can be done with cheap
green threads and blocking (the original approach of \Shake) or using
continuation-passing style (what \Shake does currently). An alternative approach to
suspending a task is to abort it and restart it again later, at the cost
of doing additional work.

@@

\vspace{-2mm}
\subsubsection{A dirty bit}\label{sec-dirty-bit}

@@

\vspace{-2mm}
\subsubsection{Verifying traces}\label{sec-verifying-traces}

### 2018-03-17 paper/5-implementations.tex
Minor tweaks
Proof S5
Add authors, minor changes

In both of the above cases, we notify the parent \hs{reordering} function of
the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
While not required for \Excel, we have implemented build systems using
\hs{reordering} which use verifying and constructive traces, effectively turning
\Excel into a cloud build system and ensuring \hs{reordering} is not overly
fitted to \Excel alone (additional implementations will be submitted to the
Artefact Evaluation).

@@

\subsection{\Shake}\label{sec-implementation-shake}

@@

\subsection{\Bazel}\label{sec-implementation-bazel}

@@

The program presented above captures certain aspects of \Bazel, but the real implementation makes one important additional assumption on \hs{Task} -- namely that it is \textit{deterministic}. With that assumption \Bazel is able to create the result hash in a trace by hashing together the result hashes of the dependencies and the key -- as the mapping between dependencies and results is deterministic. As a consequence \Bazel can compute the results of traces locally, without looking at \hs{Traces} (potentially saving a roundtrip to the server). To model this change in the code would require storing an additional transient piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)}, storing the computed hashes so far this run. With that available, \hs{getHash}~\hs{key}~\hs{s} would become:

\begin{minted}[xleftmargin=10pt]{haskell}
hash (key, [ done Map.! d | d <- ds])
\end{minted}

\noindent
And that new hash would have to be stored in \hs{done}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
cloudShake :: (Eq k, Hashable v) => Build Monad (Traces k v) k v
cloudShake = recursive $ \key fetch act -> do
    s <- gets fst
    let Traces traces contents = getInfo s
    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
    if null poss then do
        (v, ds) <- act
        modify $ \(s,done) ->
            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
                ts = Traces (t : traces) (Map.insert (hash v) v contents)
            in (putInfo ts (putValue key v s), done)
    else do
        s <- gets fst
        when (getHash key s `notElem` poss) $
            modify $ \(s, done) -> (putValue key (contents Map.! head poss) s, done)
\end{minted}
\vspace{-2mm}
\caption{An implementation of Cloud \Shake using our framework.}\label{fig-cloudshake-implementation}
\vspace{-2mm}
\end{figure}

@@

\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}

>>>>

In both of the above cases, we notify the parent \hs{reordering} function of
the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
While not required for \Excel, we have implemented build systems using
\hs{reordering} which use verifying and constructive traces, effectively turning
\Excel into a cloud build system and ensuring \hs{reordering} is not overly
fitted to \Excel alone.
%  (additional implementations will be submitted to the
% Artefact Evaluation).

@@

\vspace{-2mm}
\subsection{\Shake}\label{sec-implementation-shake}
\vspace{-1mm}

@@

\vspace{-2mm}
\subsection{\Bazel}\label{sec-implementation-bazel}
\vspace{-1mm}

@@

The program presented above captures certain aspects of \Bazel, but the real implementation makes one important additional assumption on \hs{Task} -- namely that it is \textit{deterministic}. With that assumption \Bazel is able to create the result hash in a trace by hashing together the result hashes of the dependencies and the key -- as the mapping between dependencies and results is deterministic. As a consequence \Bazel can compute the results of traces locally, without looking at \hs{Traces} (potentially saving a roundtrip to the server). To model this change in the code would require storing an additional transient piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)}, storing the computed hashes so far this run. With that available, \hs{getHash}~\hs{key}~\hs{s} would become:
\begin{minted}[xleftmargin=10pt]{haskell}
hash (key, [ done Map.! d | d <- ds ])
\end{minted}
And that new hash would have to be stored in \hs{done}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
cloudShake :: (Eq k, Hashable v) => Build Monad (Traces k v) k v
cloudShake = recursive $ \key fetch act -> do
    s <- gets fst
    let Traces traces contents = getInfo s
    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
    if null poss then do
        (v, ds) <- act
        modify $ \(s,done) ->
            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
                ts = Traces (t : traces) (Map.insert (hash v) v contents)
            in (putInfo ts (putValue key v s), done)
    else do
        s <- gets fst
        when (getHash key s `notElem` poss) $
            modify $ \(s, done) -> (putValue key (contents Map.! head poss) s, done)
\end{minted}
\vspace{-2mm}
\caption{An implementation of Cloud \Shake using our framework.}\label{fig-cloudshake-implementation}
\vspace{-4mm}
\end{figure}

@@

\vspace{-1mm}
\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
\vspace{-1mm}
